\ZpL: a p-adic precision package

### 2018-01-29 ZpL/ZpL.tex
Precision on a subset of elements and diffused digits

The lemma is more precise in the context of so-called
integral polynomial functions.
A function $f : E \to F$ is said to be
\emph{integral polynomial} if it is given by multivariate polynomial 
functions with coefficients in $\O_K$ in any (equivalently all) system 
of coordinates associated to a $\OK$-basis of $B_E(1)$.

\begin{prop}
\label{prop:precision2}
We keep the notation of Proposition \ref{prop:precision} and assume 
in addition that $f$ is integral polynomial. Let $C$ be a positive real
number such that $B_F(1) \subset df_{v_0}(B_E(C))$. 
Then Proposition \ref{prop:precision} holds with $\delta = C \cdot
\rho^{-1}$.
\end{prop}

We will use the following definition to measure the gap
between lattice and coordinate-wise methods. 
Suppose $E$ is equipped with a basis $(e_1, \ldots, e_n)$ and write $\pi_i : E \to Ke_i$ for the projections.

\begin{deftn} \label{def:diffused}
Let $H \subset E$ be a lattice. The number of 
\emph{diffused digits of precision} of $H$ is the length of
$H_0/H$ where $H_0 = \pi_1(H) \oplus \cdots \oplus \pi_n(H)$.
\end{deftn}
If $H$ represents the actual precision on some object, then
$H_0$ is the smallest diagonal lattice containing $H$.
Coordinate-wise methods, with one precision datum per coordinate,
corresponds to diagonal lattices.
Hence, they cannot yield a precision better than $H_0$, and
then $k$ provides a lower bound on the number of $p$-adic digits
gained by lattice methods over standard methods.

@@

Both \ZpLC and \ZpLF compute the absolute precision of \ttv (at time 
$t$) as the smallest valuation of an entry of the column of $M_t$ 
corresponding to the variable \ttv. Alternatively, it can be defined as 
the unique integer $N$ for which $\pi_\ttv(H_t) = p^N \Zp$ where 
$\pi_\ttv : E_t \to \Qp$ takes a vector to its \ttv-coordinate.
This definition of the absolute precision sounds revelant because, if we 
believe that the submodule $H_t \subset E_t$ is supposed to encode the 
precision on the variables in $\calV_t$, 
Proposition~\ref{prop:precision} applied with the mapping $\pi_\ttv$ 
indicates that a good candidate for the precision on $e_\ttv$ is 
$\pi_\ttv(H_t) = p^N \Zp$.

@@

\noindent \textbf{Precision on a subset of elements.}

@@

\noindent \textit{Diffused digits of precision.}

>>>>

%The lemma is more precise in the context of so-called
%integral polynomial functions.
%A function $f : E \to F$ is said to be
%\emph{integral polynomial} if it is given by multivariate polynomial 
%functions with coefficients in $\O_K$ in any (equivalently all) system 
%of coordinates associated to a $\OK$-basis of $B_E(1)$.
%
%\begin{prop}
%\label{prop:precision2}
%We keep the notation of Proposition \ref{prop:precision} and assume 
%in addition that $f$ is integral polynomial. Let $C$ be a positive real
%number such that $B_F(1) \subset df_{v_0}(B_E(C))$. 
%Then Proposition \ref{prop:precision} holds with $\delta = C \cdot
%\rho^{-1}$.
%\end{prop}

@@

\noindent
Both \ZpLC and \ZpLF compute the absolute precision of \ttv (at time 
$t$) as the smallest valuation of an entry of the column of $M_t$ 
corresponding to the variable \ttv. Alternatively, it can be defined as 
the unique integer $N$ for which $\pi_\ttv(H_t) = p^N \Zp$ where 
$\pi_\ttv : E_t \to \Qp$ takes a vector to its \ttv-coordinate.
This definition of the absolute precision sounds revelant because, if we 
believe that the submodule $H_t \subset E_t$ is supposed to encode the 
precision on the variables in $\calV_t$, 
Proposition~\ref{prop:precision} applied with the mapping $\pi_\ttv$ 
indicates that a good candidate for the precision on $e_\ttv$ is 
$\pi_\ttv(H_t) = p^N \Zp$.

@@

\noindent \textbf{Precision on a subset of elements.}
%
Our package implements the method \verb?precision_lattice? through which 
we can have access to the joint precision on a set of variables: it 
outputs a matrix (in echelon form) whose rows generates a lattice 
representing the precision on the subset of given variables.

When the variables are ``independant'', the precision lattice is
split and the method \verb?precision_lattice? outputs a diagonal
matrix:

\noindent
\begin{tabular}{rl}
\cIn  & \verb?R = ZpLC(2, print_mode='digits')? \\
      & \verb?x = R(987,10); y = R(21,5)? \\
      & \verb?R.precision().precision_lattice([x,y])? \\
\cOut & \verb?[1024    0]? \\
      & \verb?[   0   32]?
\end{tabular}

\noindent
However, after some computations, the precision matrix evolves and 
does not remain diagonal in general (though it is always triangular
because it is displayed in row-echelon form):

\noindent
\begin{tabular}{rl}
\cIn  & \verb?u, v = x+y, x-y? \\
      & \verb?R.precision().precision_lattice([u,v])? \\
\cOut & \verb?[  32 2016]? \\
      & \verb?[   0 2048]?
\end{tabular}

\noindent
The fact that the precision matrix is no longer diagonal indicates
that some well-chosen linear combinations of $u$ and $v$ are known
with more digits than $u$ and $v$ themselves. In this particular
example, the sum $u+v$ is known at precision $O(2^{11})$ while the
(optimal) precision on $u$ and $v$ separately is only $O(2^5)$.

\noindent
\begin{tabular}{rl}
\cIn  & \verb?u? \\
\cOut & \verb?...10000? \\
\cIn  & \verb?v? \\
\cOut & \verb?...00110? \\
\cIn  & \verb?u + v? \\
\cOut & \verb?...11110110110? \\
\end{tabular}

@@

\noindent \textit{Diffused digits of precision.}
%
The phenomenon observed above is formalized by the notion of
diffused digits of precision introduced in \cite{}. We recall
briefly its definition.

\begin{deftn}
\label{def:diffused}
Let $E$ be a $\Qp$-vector space endowed with a distinguished basis 
$(e_1, \ldots, e_n)$ and write $\pi_i : E \to \Qp e_i$ for the 
projections.
Let $H \subset E$ be a lattice. The number of \emph{diffused digits of 
precision} of $H$ is the length of $H_0/H$ where $H_0 = \pi_1(H) \oplus 
\cdots \oplus \pi_n(H)$.
\end{deftn}

If $H$ represents the actual precision on some object, then
$H_0$ is the smallest diagonal lattice containing $H$. It then
corresponds to the maximal \emph{coordinate-wise} precision we can 
reach on the set of $n$ variables corresponding to the basis $(e_1,
\ldots, e_n)$.
The diffused digits of precision only appear on linear combinations
of those variables.

The method \verb?number_of_diffused_digits? computes the number of
diffused digits of precision on a set of variables. Observe:

\noindent
\begin{tabular}{rl}
\cIn  & \verb?R.precision().number_of_diffused_digits([x,y])? \\
\cOut & \verb?0? \\
\cIn  & \verb?R.precision().number_of_diffused_digits([u,v])? \\
\cOut & \verb?6? \\
\end{tabular}

\noindent
For the last example, we recall that the relevant precision lattice 
$H$ is generated by the $2 \times 2$ matrix:
$$\left(\begin{matrix} 2^5 & 2016 \\ 0 & 2^{11} \end{matrix}\right).$$
The minimal diagonal suplattice $H_0$ is generated by the scalar matrix 
$2^5 \cdot \text{I}_2$ and contains $H$ with index $2^6$ in it. This is 
where the six digits of precision comes from.
There are easily visible here: the sum $u+v$ is known with $11$ digits,
that is exactly $6$ digits more than the summands $u$ and $v$.
